Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Revision of methodology by Lisa/Moritz
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2879 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3cf061283c286759eccefb6dbfb3453a6683138e

In many approaches to modeling circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, hence the location parameter of 
the von Mises distribtuion, is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by employing a generalized linear 
model (GLM)-type approach, i.e., by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This model was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
on covariates, not only in separate models but also in a mixed model. There, within an iterative 
optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
To obtain proper initial values it is recommended to apply two separate models where either one of 
the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
While being built on well elaborated theory this methodology and the contained 
optimization problem have shown to be very challenging in their application, mainly due to the 
complexity encountered in optimizing the log-likelihood function which is not globally concave 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, the selection of 
starting values is crucial and bears a high risk of the model not to converge if not done thoughtfully. 
In order to avoid this strong dependence on choosing well-fitting initial values \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM 
in a Bayesian framework.

%\fixme{revise literature survey with focus on transition to GLM}

The named methods and also many of their further developments 
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While they generally provide a powerful toolbox for modeling particularly smooth effects, 
it has to be considered that representing these smooth effects on the unit circle
can be challenging. Moreover, the discussed additive models may not be able to capture all 
types of effects adequately. For example, abrupt changes might not be detected properly owing 
to the enforced additive structure and model assumptions such as the type of link 
function and the form of linear combinations.

An alternative modeling approach that can deal with non-additive effects is provided
by tree-based methods. \cite{Lund:2002} introduced the concept of circular regression trees 
which was further explored by \cite{Larraondo+Inza+Lozano:2018}. This methodology
enables splits leading to subsets which contain the ``origin point'' $0$ equal to $2\pi$
and in that way allows for a proper representation of periodicity.

Following the idea of employing tree-based methods but including also the probabilistic concept
of regressing distribution parameters on covariates we propose a very flexible approach 
by applying the von Mises distribution within the methodology of distributional trees and 
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular data.
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.

As a motivating example Figure~\ref{fig:tree_ibk} displays a distributional tree applied to 
predict wind directions at the airport of Innsbruck in the following hour. Wind direction is 
a classical circular quantity and accurate forecasts are of great importance for decision-making 
processes and risk management in air traffic management. Innsbruck lies at the bottom of a 
deep valley in the Alps where topography channels wind along the west-east valley axis or along 
a tributary valley intersecting from the south. Hence, pressure gradients to which valley wind 
regimes react are considered as covariates along with other meteorological measurements at the 
airport (lagged by one hour), such as wind direction and wind speed at the airport.

Together with the tree structure itself Figure~\ref{fig:tree_ibk} also depicts the 
empirical (gray) and fitted von Mises (red) distribution of wind direction in each terminal 
node. Based on the fitted location parameters $\hat \mu$, the subgroups can be distinguished 
into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, node 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 10, 12 and 13). 
Node~7 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low estimated 
concentration $\hat \kappa$. In terms of covariates, the lagged wind direction 
(``persistence'') is mostly responsible for distinguishing the broad wind regimes 
listed above while the pressure gradients and wind speed separate between subgroups 
with high vs.\ low precision. Note that in the meteorological context wind direction 
is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

@@

A more extensive case study of circular regression trees and forests applied to 
probabilistic forecasting of wind direction at the airports of Innsbruck and Vienna 
is presented in Section~\ref{sec:wind} where their performances are benchmarked to 
alternative commonly used approaches.  Beforehand, distributional models for circular 
data are discussed in Section~\ref{sec:prob_circ} and the methodology of circular 
regression trees and forest and its features are presented in Section~\ref{sec:tree_forest}.
After the case study, a comprehensive summary and conclusion is given in
Section~\ref{sec:summary}.

\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
\fixme{check domain of $\mu$ in literature}
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. While applying the mixed model 
requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
However, an adaption of this methodology to circular data can be very complex due to a
challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
Moreover, difficulties might arise in case of non-additive changes and specifying a model 
properly can be a very tough task, particularly for a high number of covariates and no information
on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
employed for probabilistic forecasting of wind directions at the airport of Innsbruck. As explained
and also supported by the results illustrated in the terminal nodes of the tree in 
Figure~\ref{fig:tree_ibk} the geographic location of Innsbruck and particularly the surrounding
mountain chains are a strongly influential component of the behavior of wind directions.
While this provides an interesting and challenging environment for testing different
statistical models it has to be considered as a rather unusual location for an airport.
Therefore, in order to allow for a broad and more general evaluation and also to test
for a high flexibility of the applied methods a second site with quite converse geographic 
features is included in this study as well.
Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
area, is considered.
For the two locations the above introduced methodology of circular distributional trees and forests is
employed to wind direction data and additional meteorological measurements in order to obtain hourly 
wind direction forecasts, separately for Innsbruck and Vienna.

\subsection{Data}\label{sec:wind:data}
For the case study, as circular response variables we employ $10$\,min mean
wind direction measurements at the airports Innsbruck and Vienna on an hourly
temporal resolution. As predictor variables, we use $1$-hourly resolved
$10$\,min mean observations of various meteorological quantities such as wind
direction, wind speed, temperature, air pressure and humidity, all
lagged by one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
the vicinity of the respective airports. For Innsbruck, measurements of
21~different locations at the airport and along the intersecting valleys are
used, whereas, for Vienna, measurements of 22~locations at the airport and
within an area of about $10$\,km$^2$ are used. A topographical overview of the
airports and their surrounding areas with the station sites employed in this
study is provided in Figure~XXX. In addition, we use derived quantities such
as 3~hourly means, minima and maxima, as well as 1~and 3~hourly temporal
changes and spatial differences towards the airport of the respective
quantities. A summary of all predictor variables can be found in
Table~\ref{tab:covariates}.

The data used in this study consists of 5~total years from January 2015 to
December 2018. After first eliminating predictor variables with more than 5\% missing
values and then cases with missing observations, the
data set consists of 41979 time points and 260 covariates for Innsbruck, and of
38985 times points and 494 covariates for Vienna, respectively.

@@

\paragraph{Climatological model} 
Accurate knowledge of weather quantities' climatologies can be important for a
wide range of applications and are often also used as a baseline for the
validation of newly developed forecasting systems
\citep{simon.etal:2017,stauffer.etal:2017}. Usually, the performance of
climatological models strongly depends on the temporal dependency of the
quantity itself, either on a daily or yearly basis. Wind direction at a single
location is often quite unstable and shows only a daily or annual
seasonality in case of predominant local wind systems like the ones at
Innsbruck introduced for the analysis of Figure~\ref{fig:tree_ibk}. Hence,
the climatological model in this study provides a very simple implementation of
a baseline model and might experience serious performance losses in case of
random temporal variabilities or abrupt changes in the employed response
variable wind direction.

For simplicity, rather to fit one single climatological model, we perform the
model fitting for each time point separately with varying training data. For
each time point, the training data set consists of the same daytime as the time
of interest within 31~days centered around the same calender day over all years
except the year of interest. To gain full probabilistic climatological
forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
the distribution parameters of the von Mises distribution via maximum
likelihood estimation. The construction of a climatological reference model
was similarly pursued by \citet{vogel.etal:2018} and is discussed in a
comprehensive summary on different time-adaptive training scheme in
\citet{lang.etal:2019c}.

\paragraph{Persistency model} 
Persistence describes the previous value of a single weather quantity in a time
series; hence, it is another objective measure in the verification of
weather forecasts and also often applied as a reference model
\citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
atmospheric conditions, persistence can be a very good estimate, but as soon as
any major change takes place in the response variable during the forecast
period, forecasts based on persistence may also fail remarkably. Therefore, it
heavily depends on the variability of the variable of interest and the length
of the forecasting step whether persistence can provide overall reliable
predictions.
%If a certain weather quantity remains constant over a certain time period,
%persistence can be a very good estimate.

To gain a full probabilistic persistency model, we use as for the
climatological model maximimum likelihood estimation and fit the distribution
parameters of the von Mises distribution conditional on lagged response values
according to the description in Section~\ref{sec:prob_circ}. We fit one model
for every time point in the validation data set employing the previous six
lagged response values as training data. In order to allow for a stronger
influence of observations closer to the time of interest an exponential
smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
prediction the current observation and all previous five observations together
have an equal influence rate of 50 percent. Observations with longer time lags
have exponential weights below $0.01$ and are therefore omitted
from the training data.

\paragraph{Generalized linear model}
Traditional approaches to forecast circular response variables are often based on
circular GLMs \citep{Fisher:1993}. As discussed in
Section~\ref{sec:introduction}, circular regression models often experience
the problem that the likelihood function can be strongly irregular which makes
optimization rather difficult. Hence, they often do not converge if no
appropriate initial values are provided \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}. In this study, to be able to employ a GLM out of the box
as a reference, we use the Bayesian implementation of
\cite{Mulder+Klugkist:2017} which does not dependent so much on initial values
due to an MCMC sampling algorithm using weakly- or non-informative priors.

Following \cite{Mulder+Klugkist:2017} the employed model plugs in an
appropriate link function~$g$ in order to guarantee response values on the
interval $[-\pi,\pi]$. As the implementation cannot handle circular covariates,
we use the components of the lagged 2-dimensional wind vector $vec$ and the lagged
wind speed $spd$ as predictor variables. The model formula for the location parameter
$mu$ of the von Mises distribution can be written as:
%%
\begin{align}
\mu = \beta_0 + g(\beta_1 \cdot vec_{1} + \beta_2 \cdot vec_{2} + \beta_3 \cdot spd
\end{align}
%%
with $\beta_0$ being a circular intercept, $\beta_\bullet$ the regression
coefficients and $g$ the link function $g(x) = 2 \cdot \arctan(x)$. In addition, a
constant concentration parameter $\kappa$ is fitted to the full learning sample.

\paragraph{Circular regression tree}
For the circular regression tree introduced in Section~\ref{sec:circtree}, all
covariates provided in the learning data, as described in
Section~\ref{sec:wind:data}, can be considered due to an intrinsic automatic
variable selection performed in the tree estimation. The tree building is
performed by our newly developed R-package \pkg{circtree} employing the CTree
algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a minimal number of 2000
observations in each terminal node serving as a stopping criterion (argument
\code{minbucket}).

\paragraph{Circular regression forest} Following the description in
Section~\ref{sec:circforest}, a circular regression forest is constructed based
on 100 individual trees employing our R-package \pkg{circtree}. Each of these
trees is again built by the CTree algorithm on a subsample containing a
fraction of 30 percent of the original learning data. As it can be expected
that the lagged response variable is one of the most informative predictor
variables for wind direction forecasting, all covariates are considered for
building each tree. This `bagging' ensures that the lagged response
variable is included in each tree; in our implementation this can be done by
setting the argument \code{mtry} to the total number of covariates. Since a
high number of possible split points leads to high computational costs, the
covariates are automatically binned in a maximum of $50$ classes (argument
\code{nmax = c("yx" = Inf, "z" = 50)}). Contrary to building a single tree,
forests usually consists of large trees as they are not prone to overfitting
the data after some kind of averaging over the individual trees. Therefore, we
use the following control arguments to build rather long trees: The minimal number 
of observations to perform a split is set to 20
(argument \texttt{minsplit}), the minimal number of observations in each
terminal node is set to 7 (argument \texttt{minbucket}), and the significance
level for variable selection is kept at its maximum value of $1$
(argument \texttt{alpha}).

@@

In addition to the raw CRPS, CRPS-based skill scores are computed to assess
differences in the improvement of the various statistical models over the
climatological model used as a reference:
\begin{equation} 
  \text{CRPSS}_{\text{method}} = 1 -
\frac{\text{CRPS}_{\text{method}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

>>>>

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, i.e., the location parameter of 
the von Mises distribtuion is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This generalized linear model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
on covariates, not only in separate models but also in a mixed model. 
%There, within an iterative 
%optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
%To obtain proper initial values it is recommended to apply two separate models where either one of 
%the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
While being built on well elaborated theory this methodology 
%and the contained optimization problem have 
has shown to be very challenging in their application, mainly due to the 
complexity encountered in optimizing the log-likelihood function which is not globally concave 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, 
highly informative starting values are crucial for the model to converge at all.
In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM 
in a Bayesian framework. 
%\fixme{revise literature survey with focus on transition to GLM}
The named methods and also many of their further developments \citep[see,
e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While generally providing a powerful toolbox for modeling particularly smooth
effects, they may not be able to capture all types of effects adequately.
Especially abrupt changes might not be detected properly owing to the enforced
additive structure and model assumptions such as a possibly highly non-linear link function and
the form of linear combinations.

An alternative modeling approach that can deal with non-additive effects is provided
by tree-based methods. \cite{Lund:2002} introduced the concept of circular regression trees 
which was further extended by \cite{Larraondo+Inza+Lozano:2018}. 
Following this idea of employing tree-based methods but also including the probabilistic concept
of regressing distribution parameters on covariates, we propose a very flexible approach 
by applying the von Mises distribution within the methodology of distributional trees and 
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular data.
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.

As a motivating example, a circular distributional tree is employed to
probabilistic wind direction forecasting. Wind direction is a classical
circular quantity and accurate forecasts are of great importance for
decision-making processes especially in air traffic management.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at
the airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
Topography channels wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements at the airport (lagged by one hour), such as wind direction and
wind speed at the airport.

Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the 
empirical (gray) and fitted von Mises (red) distribution of wind direction in each terminal 
node. Based on the fitted location parameters $\hat \mu$, the subgroups can be distinguished 
into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley (from
east to west, nodes 4 and 5). 
(2) Downslope winds blowing across the Alpine crest along the intersecting valley towards 
Innsbruck (from south-east to north-west, node 8). 
(3) Down-valley winds blowing in the direction of the valley mouth (from west to east, 
nodes 10, 12 and 13). 
Node~7 captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low estimated 
concentration $\hat \kappa$. In terms of covariates, the lagged wind direction 
(``persistence'') is mostly responsible for distinguishing the broad wind regimes 
listed above while the pressure gradients and wind speed separate between subgroups 
with high vs.\ low precision. Note that in the meteorological context wind direction 
is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

@@

A more extensive case study of circular regression trees and forests applied to
probabilistic forecasting of wind direction at the airports of Innsbruck and
Vienna will be presented in this paper. Moreover, their performances are
benchmarked to alternative commonly used approaches.  The theory on
probabilistic circular modeling introducing the von Mises distribution will be
discussed in Section~\ref{sec:prob_circ}. The methodology of circular
distributional trees and forests and their features are introcuded in
Section~\ref{sec:tree_forest}.  After the case study presented in
\ref{sec:wind}, a comprehensive summary and conclusion is given in
Section~\ref{sec:summary}.

\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\section{Circular regression trees and forests}
\label{sec:tree_forest}
%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
%While applying the mixed model 
%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
%a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
%However, an adaption of this methodology to circular data can be very complex due to a
%challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
However, difficulties might arise in case of non-additive changes and specifying a model 
properly can be a very challenging, particularly for a high number of covariates and no information
on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

@@

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, industry or energy
production. Thus, for example in order to direct airplanes to a safe landing,
precise knowledge of wind direction at the respective airport for the next hour(s) is highly
desirable. This section provides a comprehensive case study on probabilistic
wind direction forecasting at two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly
introduced circular distributional trees and forests, the case study employs various
alternative approaches to forecast wind direction. The study is based on
$1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
the airports.

%
%highly fluctating, intersting variable for trees,,, requires high flexibility
%many covariables, few variables have influence, hihgly non-linear non-known interactions
%airports vienna and innsbruck
%lead times 1 and 3 hours
%
% plus figures table and fixmes
%The motivating example in Section~\ref{sec:introduction} depicts a fitted distributional tree
%employed for probabilistic forecasting of wind directions at the airport of Innsbruck. As explained
%and also supported by the results illustrated in the terminal nodes of the tree in 
%Figure~\ref{fig:tree_ibk} the geographic location of Innsbruck and particularly the surrounding
%mountain chains are a strongly influential component of the behavior of wind directions.
%While this provides an interesting and challenging environment for testing different
%statistical models it has to be considered as a rather unusual location for an airport.
%Therefore, in order to allow for a broad and more general evaluation and also to test
%for a high flexibility of the applied methods a second site with quite converse geographic 
%features is included in this study as well.
%Next to the airport of Innsbruck, located in the west of Austria, in the center of the Alps,
%a second airport in Austria, the airport of Vienna, situated in the east of the country in a rather flat
%area, is considered.
%For the two locations the above introduced methodology of circular distributional trees and forests is
%employed to wind direction data and additional meteorological measurements in order to obtain hourly 
%wind direction forecasts, separately for Innsbruck and Vienna.

\subsection{Data}\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min mean
of wind direction measurements at the airports Innsbruck and Vienna on an hourly
temporal resolution. As predictor variables, we use $1$-hourly resolved
$10$\,min mean observations of various meteorological quantities such as wind
direction, wind speed, temperature, air pressure and humidity, all
lagged by one or three hours according to the respective forecasting step. The
meteorological variables are measured either directly at the airports or within
the their vicinity. For Innsbruck, measurements of
21~different locations at the airport and along the intersecting valleys are
used, whereas, for Vienna, measurements of 22~locations at the airport and
within an area of about $10$\,km$^2$ are used. A topographical overview of the
airports and their surrounding areas with the station sites employed in this
study is provided in Figure~XXX. In addition, we use derived quantities such
as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. A summary of all predictor variables can be found in
Table~\ref{tab:covariates}.

The data used in this study consists of 5~total years from January 2015 to
December 2018. After first eliminating predictor variables with more than 5\% missing
values and then time points with missing observations, the
data set consists of 41979 time points and 260 covariates for Innsbruck, and of
38985 times points and 494 covariates for Vienna, respectively.

@@

\begin{itemize}
\item \emph{Climatological model:} 
Accurate knowledge of weather quantities' climatologies can be important for a
wide range of applications and are often also used as a baseline for the
validation of newly developed forecasting systems
\citep{simon.etal:2017,stauffer.etal:2017}. Usually, the performance of
climatological models strongly depends on the temporal dependency of the
quantity itself, either on a daily or yearly basis. Wind direction at a single
location is often quite unstable and shows only a daily or annual
seasonality in case of predominant local wind systems like the ones at
Innsbruck introduced for the analysis of Figure~\ref{fig:tree_ibk}. Hence,
the climatological model in this study provides a very simple implementation of
a baseline model and might experience serious performance losses in case of
random temporal variabilities or abrupt changes in the employed response
variable wind direction.
\newline
For simplicity, rather to fit one single climatological model, we perform the
model fitting for each time point separately with varying training data. For
each time point, the training data set consists of the same daytime as the time
of interest within 31~days centered around the same calender day over all years
except the year of interest. To gain full probabilistic climatological
forecasts, we employ the steps described in Section~\ref{sec:prob_circ} and fit
the distribution parameters of the von Mises distribution via maximum
likelihood estimation. The construction of a climatological reference model
was similarly pursued by \citet{vogel.etal:2018} and is discussed in a
comprehensive summary on different time-adaptive training scheme in
\citet{lang.etal:2019c}.
%%
\item \emph{Persistency model:}
Persistence describes the previous value of a single weather quantity in a time
series. Hence, it is another objective measure in the verification of
weather forecasts and also often applied as a reference model
\citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
atmospheric conditions, persistence can be a very good estimate, but as soon as
any major change takes place in the response variable during the forecast
period, forecasts based on persistence may also fail remarkably. Therefore, it
heavily depends on the variability of the variable of interest and the length
of the forecasting step whether persistence can provide overall reliable
predictions.
%If a certain weather quantity remains constant over a certain time period,
%persistence can be a very good estimate.
\newline
To gain a full probabilistic persistency model, we proceed similarly as for the
climatological model by using maximimum likelihood estimation and fitting the distribution
parameters of the von Mises distribution conditional on lagged response values
according to the description in Section~\ref{sec:prob_circ}. We fit one model
for every time point in the validation data set employing the previous six
lagged response values as training data. In order to allow for a stronger
influence of observations closer to the time of interest exponential
smoothing is employed with a smoothing factor of $0.5$; accordingly, for every
prediction the current observation and all previous five observations together
have an equal influence rate of 50 percent. Observations with longer time lags
have exponential weights below $0.01$ and are therefore omitted
from the training data.
%%
\item \emph{Generalized linear model:}
Traditional approaches to forecast circular response variables are often based on
circular GLMs \citep{Fisher:1993}. As discussed in
Section~\ref{sec:introduction}, circular regression models often experience
the problem that the likelihood function can be strongly irregular which makes
optimization rather difficult. Hence, they often do not converge if no
appropriate initial values are provided \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}. In this study, to be able to employ a GLM out of the box
as a reference, we use the Bayesian implementation of
\cite{Mulder+Klugkist:2017} which does not dependent so much on initial values
due to an MCMC sampling algorithm using weakly informative priors.
\newline
Following \cite{Mulder+Klugkist:2017} the employed model plugs in an
appropriate link function~$g$ in order to guarantee response values on the
interval $[-\pi,\pi]$. As the implementation cannot handle circular covariates,
we use the components of the lagged 2-dimensional wind vector $vec$ and the lagged
wind speed $spd$ as predictor variables. The model formula for the location parameter
$\mu$ of the von Mises distribution can be written as:
%%
\begin{align}
\mu = \beta_0 + g(\beta_1 \cdot vec_{1} + \beta_2 \cdot vec_{2} + \beta_3 \cdot spd)
\end{align}
%%
with $\beta_0$ being a circular intercept, $\beta_\bullet$ the regression
coefficients and $g$ the link function $g(x) = 2 \cdot \arctan(x)$. In addition, a
constant concentration parameter $\kappa$ is fitted to the full learning sample.
%%
\item \emph{Circular regression tree:}
For the circular regression tree introduced in Section~\ref{sec:circtree}, all
covariates provided in the learning data, as described in
Section~\ref{sec:wind:data}, can be considered due to an intrinsic automatic
variable selection performed in the tree estimation. The tree building is
performed by the newly developed R-package \pkg{circtree} employing the CTree
algorithm \citep{Hothorn+Hornik+Zeileis:2006} with a minimal number of 2000
observations in each terminal node serving as a stopping criterion (argument
\code{minbucket}).
%%
\item \emph{Circular regression forest:} Following the description in
Section~\ref{sec:circforest}, a circular regression forest is constructed based
on 100 individual trees employing the R-package \pkg{circtree}. Each of these
trees is again built by the CTree algorithm on a subsample containing a
fraction of 30 percent of the original learning data. As it can be expected
that the lagged response variable is one of the most informative predictor
variables for wind direction forecasting, all covariates are included for
building each tree in order to ensure that the lagged response
variable is always considered. This bagging approach can be done in \pkg{circtree} implementation by
setting the argument \code{mtry} to the total number of covariates. Since a
high number of possible split points leads to high computational costs, the
covariates are automatically binned in a maximum of $50$ classes (argument
\code{nmax = c("yx" = Inf, "z" = 50)}). Contrary to building a single tree,
forests usually consists of large trees as they are not prone to overfitting
the data after some kind of averaging over the individual trees. Therefore, we
use the following control arguments to build rather large trees: The minimal number 
of observations to perform a split is set to 20
(argument \texttt{minsplit}), the minimal number of observations in each
terminal node is set to 7 (argument \texttt{minbucket}), and the significance
level for variable selection is kept at its maximum value of $1$
(argument \texttt{alpha}).
\end{itemize}

@@

In addition to the raw CRPS, also CRPS-based skill scores are computed to assess
differences in the improvement of the various statistical models over the
climatological model used as a reference:
\begin{equation} 
  \text{CRPSS}_{\text{model}} = 1 -
\frac{\text{CRPS}_{\text{model}}}{\text{CRPS}_{\text{climatology}}}.
\end{equation}

All scores presented in the next section are out-of-sample based on five years
of observations: For the persistency model only time points prior to the time
of interest are used and the validation is performed rolling over all
observations. For all other models a five-fold cross-validation is employed
using up to four calendar years for model training and the remaining single
calendar year for validation. Due to the large sample size of 24 hourly values
per day over five years, some kind of temporal aggregation is needed to ensure
a correct visual comparison of the individual methods.  Performed analyses have
shown that the for the employed models the variability of the predictive
performance over the five years is lower than over a single day or over a
single year. Hence, CRPS and CRPS skill scores are aggregated over the
respective five validation years which yields 24 hourly scores per month
averaged over the five validation years.
